Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-02-18 Paper/Paper.tex
Updates from Overleaf

\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
\usepackage{setspace}
\onehalfspacing
\addtolength{\skip\footins}{2pc plus 5pt}

@@

The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making. Their interaction is relevant in several aspects: technology causing displacement, whether low-skilled vs. high-skilled workers are differentially impacted, or the productivity gains that can be realized due to technology. In many capacities, across differing countries and industries, people tend to care about how technology affects current and future labor markets. This paper seeks to better understand this relationship under a major technology change in the U.S. healthcare system, the implementation of electronic health care record systems in hospitals.

Electronic Health Records (EHRs) have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on cost and quality of care. For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Improvement in quality of care was expected as a result of physicians having more time to spend with patients, better communication across different physicians on the same care plan, and decision-making assistance to avoid harmful mistakes. Such expectations about cost and quality led the government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech}). This legislation subsidized hospitals which used EHRs “meaningfully”\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The movement towards EHR use is evidenced in hospitals from 2008 onward; the percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}).

Specifically, an EHR is digital version of a patient’s medical records. These include detailed accounts and notes of medical history and can become advanced enough to provide the physician suggestions for care. Physicians are the primary users of this technology and have a significant role in determining whether potential benefits of EHRs are realized. However, the day-to-day life of physicians changed drastically with the rapid implementation of EHRs. What was expected to be a helpful tool turned out to be burdensome and frustrating for the persons using it. In individual interviews, multiple physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). A physician with the choice to retire or move to a smaller private practice may choose to do so when the cost of using an EHR in a hospital becomes too high. This paper seeks to understand whether the implementation of EHRs in hospitals led to changes in physician labor market outcomes: where to work, productivity levels, and retirement.

Using CMS Shared Patient Data from 2009-2015, I construct a panel of physician-hospital pairs that captures a specific working relationship between the pair, most likely the physician working in the hospital. I link the hospitals to AHA survey data for information on EHR use and then aggregate the data to the physician level. The key independent variable for my main analysis is an individuals physicians' exposure to electronic health records. There are three outcomes of interest in this paper. First, I consider the decision for a physician to retire. \textcolor{red}{This is measured using blank}. Second, I consider whether those physicians who stay in the labor force choose to substitute work towards office-based practices. \textcolor{red}{This information comes from blank.} Finally, for physicians who continue working in hospitals, I analyze the productivity impacts of EHRs. The measure of productivity used is the number of Medicaid patients that bill the both the physician and a hospital in the same day, which comes from the CMS Shared Patient Data.

The estimation is an event-study approach, where treatment is whether a physician is exposed to an EHR. I assume the decision of a hospital to implement an EHR is exogenous to the physician, but explore this assumption in more detail in Section \textcolor{red}{blank}. \textcolor{red}{To extend the analysis I also utilize difference in difference with continuous treatment, using fraction of hospitals using an EHR as the treatment variable.}

\textcolor{red}{Paragraph about findings}

This paper contributes to the general understanding of two main questions: the broad effects of technology on labor markets, and the more specific effects of EHR implementation in various healthcare dimensions.

How technology affects labor market outcomes depends drastically on the industry and setting being considered. Health is a major component of expense in the United States, accounting for approximately 17 percent of GDP. Within the health sector, health information technology has been a major change implemented widely. There is a continual discussion amongst policymakers and researchers about how to lower healthcare spending without sacrificing quality of care. Thus, the potential cost savings of EHRs are extremely important to understand given their potential to also improve patient satisfaction, since there is a continual interest in achieving these two things simultaneously. Further, there is the question in both economics and medical literature of whether there exists a physician shortage and, similarly, access to care issues in the U.S. (\cite{cooper2002economic}). A relevent factor of this discussion is whether electronic health records are driving physicians out of the market or decreasing the amount of patients that can be seen by one physician, both of which have the potential to worsen these issues.

The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively. Despite many case studies and hospital-level analysis that indicate large improvements in health outcomes such as mortality (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}). Further, there have been no significant cost decreases due to EHRs in the short or long run. If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}). However, the incentive programs put in place could impact the cost effects of EHRs, and 6 years may not be a long enough time frame to see the full picture of cost effects. While this paper does not specifically investigate hospital costs, it does speak to potential drivers of cost effects. Finally, there have been preliminary case studies and interviews seeking to understand physicians' outlook on EHRs. These differ depending on the setting studied, where there is evidence of older physicians having more frustration regarding the technology than younger physicians. \textcolor{red}{I need a citation here and probably to expand more on the idea of physician opinions.}

This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010, capturing more so the effect of computerization.

@@

\textcolor{red}{Need to figure out what goes in this section. History of EHRs? Specifics of the subsidies and meaningful use?}

@@

Overall summary statistics of the variables are shown in \hyperref[fig:summarystatistics]{Table 1}. Average hospital operating days is a physician-level variable that averages the number of operating days for that physicians hospital network. While there are some low outliers with a low value, on average physicians are working in hospitals open year round. Similarly, the average size of the hospital worked with averages the number of beds of each hospital in a physician's network. While these two average variables can technically vary by year, they tend to stay constant across the data, which is why they are included in the static category. Approximately 30\% of the sample of physicians are female and the average physician graduated medical school in 1993. Note that there are no physicians in the data who graduated after 2009, since they would be entering the labor force during the span of the data. On average, a physician works with between 1 and 2 hospitals as well as hospital systems. While the number of hospitals that a physician cares for patients in varies by year, the number of hospitals in their network is constant. The physician level variables that vary over time deal are the treatment variables regarding EHR exposure and the number of Medicaid patients seen. For all physician-years, physicians were exposed to an EHR in 56\% of the hospitals they work in and 70\% of the sample is exposed to any EHR. Once a physician becomes exposed to an EHR, they cannot go back to being unexposed. Finally, physicians are seeing an average of 620 Medicaid patients in hospitals per year. \textcolor{red}{In Appendix A1, I include summary statistics by subsample of physician age and year.}

\begin{table}[ht]
    \import{Table Code}{overall stats.tex}
    \label{fig:summarystatistics}
    \vspace{1mm}
    \caption*{\footnotesize Obs: 1,044,001}
\end{table}

@@

\begin{figure}
    \caption{Treatment Variables Over Time}
    \vspace{-2mm}
    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
    \label{fig:figure1}
\end{figure}

@@

One concern with this measure of productivity is that an increase in the number of patients seen by a physician could be caused by an outside factor that is also correlated with EHR adoption in hospitals. A logical outside factor could be the existence of employees with the purpose of utilizing electronic health records on a physician's behalf. Employees of this type are assigned a medical tax ID when working in hospitals; they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator. All of these employees are the equivalent of transcribers for health information technology.

Using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals. Figure \textcolor{red}{number} is a frequency plot of the year of activation for every tax code that falls in the categories listed above. The total number of these registered employees is 875.

@@

\section{Tables and Figures}

\begin{table}[ht]
\import{Table Code}{split stats.tex}
\end{table}

>>>>

\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
\usepackage{setspace}
\doublespacing
\addtolength{\skip\footins}{2pc plus 5pt}

@@

The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making. Their interaction has direct impacts across industries; job displacement, creation, and satisfaction are all prevalent reality for many workers (\textcolor{red}{PUT SOME CITATIONS HERE}). Additionally, in some cases, labor market changes that take place for one job can have rippling effects to populations of people. One such job is that of a physician, where physician decisions have direct and indirect effects on patients receiving care. For example, a physician experiencing burnout from work pressures may treat patients differently and lead to patients feeling uncomfortable seeking care (\textcolor{red}{FIND A CITE FOR THIS)}, or increased administrative burdens could lead to less time with patients and exacerbated access to care issues for those in rural settings (\textcolor{red}{FIND CITE FOR THIS}). There may also be positive effects on patients from physician labor market changes, such as increased access to care that comes from more physicians entering the market, or less harmful mistakes that arise from standardizing care given. For these reasons, I investigate physician labor markets under a major technology change in the U.S. healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce labor market changes in the occupation. \textcolor{red}{Rethink how I frame the examples, but MAKE SURE I HAVE CITATIONS FOR THESE THINGS}

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggesions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency of care. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), evidence of the nationwide movement towards digitization.

Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized. However, with the rapid implementation of EHRs, the day to day life of physicians changed drastically. In interviews, groups of physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices. Further, the short and long run impacts of EHRs on physician productivity could vary greatly.  I use modern event study approaches, including continuous treatment, to understand the impact of EHR implementation in hospitals on physician labor market outcomes: retirement, where to work, and productivity in hospitals.

Using CMS shared patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct physician-level data on hospitalists and other general practice physicians connected to hospitals from 2009-2017. The data captures the extent to which the physician in exposed to EHRs in hospitals and various labor market outcomes, along with demographic information such as zip code, age, and gender. With this data I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician. Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals.  For retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it. Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here (if no effect, take out the first part of this sentence)}.

Given that there is a population of both younger and older physicians who stay in the labor force (\textcolor{red}{it may be that all of them stay in the market}), the next outcome to consider is whether physicians switch work settings as a result of EHR implementation in hospitals. That is, do physicians switch from primarily hospital based work to smaller office based settings, or do they switch hospital locations where the technology looks different?  Here, I limit the sample to any physicians who stay in the labor force after EHR exposure. I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.

These findings imply that there are still physicians who remain employed by the same hospital after EHR implementation. Thus, I study the effect of the technology on productivity within hospitals, where productivity is measured by the number of patients seen. The population of physicians left in this analysis are those who do not retire or switch to an office based setting. I find that productivity does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains. After 5 years of exposure, a physician is seeing on average 11\% more patients relative to before EHRs. This result speaks to the disconnect between expected effects and realized effects of EHRs which is discussed in the health information technology literature. Since more patients are being seen by physicians, efficiency and access to care have both increased while costs (Medicaid patients) and quality remain unchanged. \textcolor{red}{This paragraph needs some work as my results have changed since getting the MDPPAS data and I need to push on other mechanisms besides productivity - if other physicians left and work had to be picked up or billing activity increased or demand increased (maybe due to ACA?)}

\textcolor{red}{Add a paragraph discussing sensitivity analysis or mechanisms.}

A key assumption underlying this analysis is that EHR exposure is exogenous to individual physicians. That is, a hospital makes the decision to implement an EHR independent of a physicians' labor market response to that implementation. This assumption is not unrealistic in large hospitals who employ many physicians, but may be violated in small hospitals where a physician has a voice in management. To address this, I first limit the analysis to large hospitals. However, small hospitals are especially important in studying patient access to care. To include small hospitals in the analysis, I instrument EHR adoption with local bandwidth capacity. \textcolor{red}{Relate these findings to previous findings. I need to make this paragraph more clear as I improve what I'm doing}

This paper's contribution to the space is threefold: (1) this is the first study to my knowledge that empirically analyzes the relationship between physician labor markets and EHRs, (2) since there seems to be a disconnect between the potentials of EHR technologies and the realized effects, this paper will speak to a potential mechanism that is causing this puzzle, and (3) this study covers the time period of the major EHR boom in healthcare, whereas most studies mentioned above utilize data before 2010. \textcolor{red}{This needs to be expounded upon a lot. Perhaps one or two paragraphs for each area of contribution.}

The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively. Despite many case studies and hospital-level analysis that indicate s large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}). Further, there have been no significant cost decreases due to EHRs in the short or long run. If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}). As discussed earlier, these findings are puzzling due to the expected benefits of EHRs. This paper contributes to our knowledge of this area in re-framing past finding into a patient access to care problem instead of an efficiency problem.

In 2009, the

@@

EHRs have been an important feature of healthcare since the 1980s and 1990s. Early on, this technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling. Before the rapid development of computers, physicians were not even able to interface directly with EHRs, and thus were not drastically affected. The cost of utilizing an EHR was extremely large and increasing in the amount of capabilities, which led to the widespread view that they were only complements to paper records and not feasible as a substitute. The feasibility of using only EHRs increased as computers became more portable, creating what is known as the "physician workspace", a computer station for the physician to interface directly with EHRs. However, physicians kept the view that the technology was purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user. Even when automation was developed for certain data, policy required users to double check all of the data in the EHR for errors.

The HITECH Act was passed in 2009, setting aside government funds to subsidies hospitals who used EHRs according to certain guidelines. These requirements include having at least 80\% of patients in the system, asking certain questions to patients, and protecting the security of the system. Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies. Hospitals began receiving payments for the first stage of meaningful use in 2011. This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). \textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, some physicians still continue to report frustration. "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}). This shows the continued hesitancy to utilize EHRs that physicians exhibit.

@@

Overall summary statistics of the variables are shown in \hyperref[fig:summarystatistics]{Table 1}. Average hospital operating days is a physician-level variable that averages the number of operating days for that physicians hospital network. While there are some low outliers with a low value, on average physicians are working in hospitals open year round. Similarly, the average size of the hospital worked with averages the number of beds of each hospital in a physician's network. While these two average variables can technically vary by year, they tend to stay constant across the data, which is why they are included in the static category. Approximately 30\% of the sample of physicians are female and the average physician graduated medical school in 1993. Note that there are no physicians in the data who graduated after 2009, since they would be entering the labor force during the span of the data. On average, a physician works with between 1 and 2 hospitals as well as hospital systems. While the number of hospitals that a physician cares for patients in varies by year, the number of hospitals in their network is constant. The physician level variables that vary over time deal are the treatment variables regarding EHR exposure and the number of Medicaid patients seen. For all physician-years, physicians were exposed to an EHR in 56\% of the hospitals they work in and 70\% of the sample is exposed to any EHR. Once a physician becomes exposed to an EHR, they cannot go back to being unexposed. Finally, physicians are seeing an average of 620 Medicaid patients in hospitals per year.

\vspace{5mm}
\begin{table}[ht]
    \import{Table Code}{overall stats.tex}
    \label{fig:summarystatistics}
    \vspace{1mm}
    \caption*{\footnotesize Obs: 1,044,001}
\end{table}
\vspace{5mm}

I also include summary statistics by year and physician experience for comparison in Table \textcolor{red}{number}. In all years, senior physicians see more patients on average than younger physicians. Note that in 2015 the number of Medicaid patients drops drastically. This is due to the data collection in 2015 cutting off earlier than previous years since it was the last year collected for the data set. Because of this limitation, I drop 2015 in the productivity analysis, where number of patients is the dependent variable.

\textcolor{red}{I need to think about why senior physicians are more exposed than younger physicians}

\vspace{5mm}
\begin{table}[ht]
\caption{Summary Statistics by Year and Experience Level}
\import{Table Code}{split stats.tex}
\end{table}
\vspace{5mm}

@@

\vspace{5mm}
\begin{figure}[ht]
    \caption{Treatment Variables Over Time}
    \vspace{-2mm}
    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
    \label{fig:figure1}
\end{figure}
\vspace{5mm}

@@

One concern with this measure of productivity is that an increase in the number of patients seen by a physician could be caused by an outside factor that is also correlated with EHR adoption in hospitals. A logical outside factor could be the existence of employees with the purpose of utilizing electronic health records on a physician's behalf. Employees of this type are assigned a medical tax ID when working in hospitals; they have the following titles: Coding Specialist (Hospital Based), Health Information Technologist and Registered Record Administrator. I will refer to any employee in these categories as a data assistant.

Using NPPES data on all NPIs and their tax information, I analyze the existence of data assistants in hospitals. Figure \textcolor{red}{number} is a frequency plot of the year of activation for every tax code that falls in the categories listed above. The total number of these registered employees is 875. The first data assistants ever registered were in 2005, when electronic medical records were being implemented. From 2005 to 2013, more 15-60 more employees were registered each year. The graph is clearly skewed towards later years, where a significant increase in the number of new data assistants occured in 2014. If hiring data assistants were directly correlated with both EHR implementation and physician frustration, one would expect the increase to occur from 2011-2012, when a majority of physicians were first exposed to EHRs.

\vspace{5mm}
\begin{figure}[ht]
\caption{Frequency of Data Assistant Enumeration by Year}
    \includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
    \label{fig:dataassistant_histogram}
\end{figure}

The total number of registered data assistants may not equal the true number of people doing a data assistant job in hospitals. However, under the assumption that the number of registered data assistants is proportional to the number of actual data assistants, the claim of no correlation still holds.

@@

\section{}
